---
title: "Orthology Search"
author: "Fiona Messer"
date: "2026-09-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

Converting a long list of genes you are interested in from *Drosophila pseudoobscura* to *D. melanogaster* is time-consuming and prone to error. With this script, you can supply lists of genes, either manually in the script or as a csv file, and it will convert the whole list in one go.

Furthermore, you can supply genes from both *D. pseudoobscura* and *D. melanogaster* in the same list. The supported formats are CG number, LOC number and gene symbol. Inputs are not case sensitive.

This script requires the R package `stringr`.

## Inputs

Provide a list of genes you want to convert. You can provide a list in the script, with each gene in quotes and separated by a comma (e.g. `"cg3399", "Act42a", "4805701"`). Alternatively, you can supply a csv file. Simply copy the file into the R project folder, then provide the name of the csv file in the `markers` field (e.g. `"genes_of_interest"`). You do not need to add the file extension.

```{r marker input}
markers <- c("cg3399", "Act42a", "4805701") # List of genes or name of csv file
```

Next, let the script know whether you have provided a list or a csv file.

```{r file type}
file_type <- "list" # type of file provided, either "list" or "csv"
```

Now decide how you would like your outputs. You can save the output to a csv file, and if so, add a specific prefix to your output file.

```{r outputs}
save_orthologs <- FALSE # Logical (TRUE or FALSE) TRUE exports converted genes to .csv file

prefix <- "" # Add a prefix to your output file names
```

That's it for inputs, now the script will run and generate the list of orthologous genes.

## Orthology search script

The script loads or imports the database of orthologous genes.

```{r load database, echo=FALSE}
# Load the database
if (exists("dpse_orthology") && is.data.frame(get("dpse_orthology"))){
  print("database pre-loaded")
} else {
  dpse_orthology <- read.csv("dpse_dmel_orthology.csv")
  print("database imported")
}
```

Next it loads the list of genes provided by the user.

```{r marker list, echo=FALSE}
# Load the marker list
if (file_type == "csv"){
  markers <- read.csv(paste0(markers, ".csv")) # Loads the list of genes
  print("markers provided as csv")
} else {
  print("markers provided as list")
}
```

Remove any data from previous searches:

```{r clear previous, echo=FALSE}
# clear outputs from previous searches
matching_values <- NULL
orthologs <- NULL
search_column <- NULL
```

And run the search:

```{r stringr search, echo=FALSE}
# search with stringr
for (m in markers){
  if((grepl("CG", m, ignore.case = TRUE)) == TRUE){
    search_column <- dpse_orthology$Flybase.annotation.ID
    print("gene type = CG")
  } else if((grepl("LOC", m, ignore.case = TRUE)) == TRUE){
    search_column <- dpse_orthology$Dpse.Gene.stable.ID
    print("gene type = LOC")
  } else if((grepl("FBgn", m, ignore.case = TRUE)) == TRUE){
    search_column <- dpse_orthology$Dmel.Gene.stable.ID
    print("gene type = FBgn")
  } else {
    search_column <- NULL
    print("gene type = unknown") #tells stringr which column to serach depending on type of input
  }
  if(length(search_column) < 1){
    matching_values <- append(matching_values, 
                              str_which(dpse_orthology$LOC, 
                                        regex(paste0("^", m, "$"), ignore_case = TRUE)))
    matching_values <- append(matching_values, 
                              str_which(dpse_orthology$Dmel.Gene.name.y, 
                                        regex(paste0("^", m, "$"), ignore_case = TRUE)))
    matching_values <- append(matching_values, 
                              str_which(dpse_orthology$Dmel.Gene.description, 
                                        regex(paste0("^", m, "$"), ignore_case = TRUE)))
  } else {
    matching_values <- append(matching_values, 
                              str_which(search_column, 
                                        regex(paste0("^", m, "$"), ignore_case = TRUE)))}
  print(matching_values)
}
print("search completed")
```

## Outputs

A gene orthology dataframe is generated, containing both *D. pseudoobscura* and *D. melanogaster* orthologues of all of the provided genes.

The orthology table is printed below:

```{r output dataframe, echo=FALSE}
orthologs <- dpse_orthology[matching_values,c(2, 3, 7, 11, 13)]
orthologs

if(save_orthologs == TRUE){
  write.csv(orthologs, paste0(prefix, "orthologues.csv"))
  print("Output saved as csv")
} else {
  print("WARNING: Output not saved")
}
```

The end.